$Z_3$ topological order in the quantum dimer-pentamer model

### 2016-11-27 qdpm_paper.tex
small changes to paper

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_3way.pdf}
        \caption
        {
            (a) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}}
            n_{\bm{r}}\rangle$ on an $L=32$ lattice. The origin corresponds to the orange link, and
            $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off
            the given scale. 
            (b) The horizontal dimer density structure factor $S_d(\bm{k})$ for the
            QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational
            symmetry is due to the fact that horizontal links have been isolated. 
            (c) The horizontal dimer-dimer structure factor $S_d(\bm{k})$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:dimer_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($|S_M|$) structure factors. All maximum vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. \note{Owen: We don't need the absolute value bars for $S_d$ since it's real.}}
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{vison_3way.pdf}
        \caption
        {
            (a) Color scale of the magnetic string correlation function 
            on an $L=32$ lattice. The origin corresponds to the orange circle and it has been removed from
            the color scale.
            (b) The magnitude of the $Z_3$ magnetic string correlation function
            $|C_M(\bm{r}_{\bm{0}p})|$, where the correlations is with respect to the orange plaquette.
            (c) The magnitude of the $Z_3$ magnetic string structure factor $|S_M|$ shown along a
            path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols.
        }
        \label{fig:vison_3way}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state.

The above results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions. The local constraints of the QDM can be realized in two-body spin-$1/2$ and Bose-Hubbard models, and the QDM Hamiltonian can arise perturbatively in the constrained low energy Hilbert space~\cite{Balents2002a,Zhitomirsky2005,Isakov2006c,Albuquerque2008}. Thus future work, it may be possible to engineer the QDPM in the low energy subspace of an experimentally realizable two-body model.

@@

Our calculations were performed on the VACC....

>>>>

%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{dimer_gry_scale_qdpm.pdf}
        \caption{({\it left}) Gray scale of the dimer-dimer correlation function $\langle n_{\bm{0}} n_{\bm{r}}\rangle$ on an $L=16$ lattice. The origin corresponds to the blue link, and $\bm{r}$ varies over all other links of the lattice. Red links have correlations that are off the given scale. ({\it right}) The magnitude of the $Z_3$ magnetic string correlation function $|C_M(\bm{r}_{\bm{0}p})|$, where the the correlations is with respect to the blue plaquette. \note{Owen:What if we put the dimer structure factor on the right panel here, and move the vison correlation to the left pannel of figure 6, and put the top panel of figure 7 below these two?}}
        \label{fig:dimer_vison_heatmap}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
   \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{qdpm_dmr_vis_HM.pdf}
        \caption{({\it left}) The horizontal dimer density structure factor $S_d(\bm{k})$ for the QDPM RK ground state on an $L=64$ lattice. The lack of $x\leftrightarrow y$ rotational symmetry is due to the fact that horizontal links have been isolated. ({\it right}) \note{Owen: Is this the real part or $\sqrt{S_M^* S_M}$}}
        \label{fig:structure_factor} 
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
    \begin{figure}[t]
        \centering
        \includegraphics[width=1.0\columnwidth]{vis_dmr_qdpm_struc_fac_32x32.pdf}
        \caption{ The horizontal dimer-dimer structure factor $S_d(\bm{k})$ ({\it top}) and the magnitude of the $Z_3$ magnetic string structure factor $|S_M|$ ({\it bottom})
        shown along a path in $k$ space for an $L=64$ lattice. Statistical errors are smaller than the symbols. \note{Owen: Make the k's in the y-axis bold. Switch the y-axis from $S_v\rightarrow S_M$.}}
        \label{fig:structure_factor_path}
    \end{figure}
%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \centering
    \includegraphics[width=1.0\columnwidth]{peak_scaling.pdf}
    \caption{Finite-size scaling of the maxima in the dimer ($S_d$) and magnetic string ($|S_M|$) structure factors. All maximum vanish in the thermodynamic limit, indicating that the RK state of the QDPM is a symmetric dimer liquid. \note{Owen: We don't need the absolute value bars for $S_d$ since it's real.}}
    \label{fig:bragg_scaling}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

@@

We have introduced the quantum dimer-pentamer model on the square lattice and studied the ground state at the RK point. Using numerical calculations, we have demonstrated that the RK state of the QDPM is a dimer liquid, without spontaneously broken symmetry. The extensive topological degeneracy of the square lattice QDM has been reduced to a finite 9-fold degeneracy in the QDPM on a torus. We find that the dimer correlations decay exponentially with distance in contrast to the power-law decay of dimer correlations in the QDM. Additionally we also have demonstrated that monomer defects are deconfined in the QDPM, in contrast with the critical confinement of monomers in the QDM. Finally, we find that the $Z_3$ magnetic string correlations decay exponentially and display no evidence of $Z_3$ vortex condensation; this is consistent with $Z_3$ vortices being gapped quasiparticles excitations above the QDPM ground state. These results suggest a the low energy physics of the QDPM is described by a $Z_3$ gauge theory with a $Z_3$ topological ordered ground state. Thus we have demonstrated that systems with Ising degrees of freedom and local constraints can lead to $Z_N$ topological order for $N>2$, providing the a new stable phase of matter beyond the commonly seen $Z_2$ topological spin liqud.

@@

Finally, given the stability of the topologically ordered phase, it may be possible to realize the QDPM in a more microscopically realistic spin model with two body interactions. The local constraints of the QDM can be realized in two-body spin-$1/2$ and Bose-Hubbard models, and the QDM Hamiltonian can arise perturbatively in the constrained low energy Hilbert space~\cite{Balents2002a,Zhitomirsky2005,Isakov2006c,Albuquerque2008}. Thus future work, it may be possible to engineer the QDPM in the low energy subspace of an experimentally realizable two-body model, providing a route to the experimental realization of a $Z_3$ topologically ordered spin liquid.

@@

Computations were performed on the Vermont Advanced Computing Core supported by NASA (NNX-08AO96G).
